\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}

### 2017-07-11 tex/paper.tex
connection to variational inference via posterior predictive distilled

\newcommand{\qxpsi}{q(\mathbf{x}|\bfpsi)}
\newcommand{\bftheta}{{\bm \theta}}
\newcommand{\bfpsi}{{\bm \psi}}
\newcommand{\bfx}{\mathbf{x}}
\newcommand{\bfz}{\mathbf{z}}

@@

Given some observed data $\{ \mathbf{x}_i | i=1, \dots, N \}$ drawn from the
(unknown) true distribution $p_r(\mathbf{x})$, our goal is to estimate the parameters 
$\bftheta^*$ that minimize the divergence between $p_r(\mathbf{x})$ and
the implicit model $p(\mathbf{x}|\bftheta)$. That is,
\begin{equation}
    \bftheta^* = \arg \min_{\bftheta \in \bftheta} \rho(p_r(\mathbf{x}), p(\mathbf{x}|\bftheta)),
\end{equation}
where $\rho$ is some distance or divergence.

@@

Generative adversarial networks (GANs) were first proposed by
\cite{goodfellow2014generative} as a way to build an implicit generative model
capable of producing samples from random noise $\bfz$. More specifically,
a generative model $g(\cdot; \bftheta)$ is pit against an adversarial
classifier $d(\cdot; \phi):\mathbb{R}^d \to [0,1]$ with parameters $\phi$ and whose antagonistic objective is to recognize real data $\mathbf{x}$
from generated data $\tilde{\mathbf{x}} = g(\bfz; \bftheta)$. Both models $g$ and $d$
are trained simultaneously, in such a way that $g$ learns to fool
its adversary $d$ (which happens when $g$ produces samples comparable to the
observed data), while $d$ continuously adapts to changes in $g$. When $d$ is
trained to optimality before each parameter update of the generator, it can
be shown that the original adversarial learning procedure~\cite{goodfellow2014generative} amounts to minimizing
the Jensen-Shannon divergence $\text{JSD}(p_r(\mathbf{x}) \parallel p(\mathbf{x}|\bftheta))$ between $p_r(\mathbf{x})$ and $p(\mathbf{x}|\bftheta)$.

As thoroughly explored in \citep{2017arXiv170104862A}, GANs remain remarkably
difficult to train because of vanishing gradients as $d$ saturates, or because of
unreliable updates when the training procedure is relaxed. As a remedy,
Wasserstein GANs~\citep{2017arXiv170107875A} reformulate the adversarial
setup in order to minimize the Wasserstein-1 distance $W(p_r(\mathbf{x}), p(\mathbf{x}|\bftheta))$ by
replacing the adversarial classifier with a 1-Lipschitz adversarial critic
$d(\cdot; \phi) : \mathbb{R}^d \to \mathbb{R}$. 
\begin{align}
   W( p_r(\mathbf{x} ), p(\mathbf{x} | \bftheta)  ) = \mathcal{L}_W = \sup_{f} \mathbb{E}_{p(\tilde{\mathbf{x}}|\bftheta)} [f(\tilde{\mathbf{x}})] - \mathbb{E}_{p_r(\mathbf{x})} [f(\mathbf{x})]
\end{align}
Under the WGAN-GP formulation of \cite{2017arXiv170400028G}
for stabilizing the optimization procedure,
training $d$ and $g$ results in alternating gradient updates on $\phi$ and $\bftheta$ in order to respectively minimize

\begin{align}
    {\cal L}_d =\,&  {\cal L}_W + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\phi) ||_2 - 1)^2] \\
    {\cal L}_g =\,& - {\cal L}_W
\end{align}
%\begin{align}
%    {\cal L}_d =\,& \mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)} [d(\tilde{\mathbf{x}};\phi)] - \mathbb{E}_{\mathbf{x} \sim p_r(\mathbf{x})} [d(\mathbf{x};\phi)]  \nonumber \\
%                  & + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\phi) ||_2 - 1)^2] \\
%    {\cal L}_g =\,& -\mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)} [d(\tilde{\mathbf{x}};\phi)]
%\end{align}
where ${\hat{\mathbf{x}}} := \epsilon \mathbf{x} +
(1-\epsilon)\tilde{\mathbf{x}}$, for $\epsilon \sim U[0,1]$, $\mathbf{x} \sim
p_r(\mathbf{x})$ and $\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)$.

@@

Variational optimization~\cite{2012arXiv1212.4507S,staines2013optimization} and evolution strategies~\citep{2011arXiv1106.4487W} are general
optimization techniques that can be used to form a differentiable bound
on the optima of a non-differentiable function. Given a function $f$ to minimize,
these techniques are based on the simple fact that
\begin{equation}
    \min_{\bftheta \in \bftheta} f(\bftheta) \leq \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta)] = U(\bfpsi),
\end{equation}
where $q(\bftheta|\bfpsi)$ is a proposal distribution with parameters $\bfpsi$ over input values $\bftheta$.
That is, the minimum of a set of function values is always less than or equal
to any of their average. Provided that the proposal is flexible enough, the parameters $\bfpsi$
can be updated to place its mass arbitrarily tight around the optimum $\bftheta^* = \min_{\bftheta \in \bftheta} f(\bftheta)$.

Under mild restrictions outlined in  \citep{2012arXiv1212.4507S}, the bound
$U(\bfpsi)$ is differentiable with respect to $\bfpsi$, and using the log-likelihood
trick it comes:
\begin{align}\label{eqn:approx-grad}
    \nabla_\bfpsi U(\bfpsi) &= \nabla_\bfpsi \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta)] \nonumber \\
    %&= \nabla_\bfpsi \int f(\bftheta)  q(\bftheta|\bfpsi)  d\bftheta \nonumber \\
    &= \int f(\bftheta) \nabla_\bfpsi q(\bftheta|\bfpsi)  d\bftheta \nonumber \\
    &= \int \left[ f(\bftheta) \nabla_\bfpsi \log q(\bftheta|\bfpsi) \right]  q(\bftheta|\bfpsi) d\bftheta \nonumber \\
    &= \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta) \nabla_\bfpsi \log q(\bftheta|\bfpsi)]
\end{align}
Effectively, this means that provided that the score function $\nabla_\bfpsi \log
q(\bftheta|\bfpsi)$ of the proposal is known and that one can evaluate
$f(\mathbf{\bftheta})$ for any $\bftheta$, then one can construct empirical
estimates of Eqn.~\ref{eqn:approx-grad}, which can in turn be used to minimize
$U(\bfpsi)$ with stochastic gradient descent (or a variant thereof, like
Adam~\cite{2014arXiv1412.6980K} or the Natural Evolution Strategy
algorithm~\citep{2011arXiv1106.4487W}, for scaling invariance and
robustness to noisy gradients).

% In this simple form, stochastic gradient descent based on
% Eqn~\ref{eqn:approx-grad} is known to suffer from premature convergence and lack
% of scale invariance. To avoid the dependence on the parameterization of the
% proposal distribution, natural gradients~\citep{amari1998natural} can be used
% instead to point towards the steepest descent in the space of realizable
% proposal distributions, rather than in the space of their parameters. Effectively,
% the natural evolution strategy (NES) algorithm~\citep{2011arXiv1106.4487W}
% corrects the gradient of $U$ according to the local curvature of the KL-divergence
% surface of proposal distributions, which results in forming
% \begin{equation}\label{eqn:approx-grad-nat}
%     \tilde{\nabla}_\bfpsi U(\bfpsi) = \mathbf{F}(\bfpsi)^{-1} \nabla_\bfpsi U(\bfpsi),
% \end{equation}
% where $\mathbf{F}(\bfpsi)$ is the Fisher information matrix
% \begin{equation}
%     \mathbf{F}(\bfpsi) = \mathbb{E}_{\bftheta \sim q_\phi}[ \nabla_\bfpsi \log q_\bfpsi(\bftheta) \nabla_\bfpsi \log q_\bfpsi(\bftheta)^{\top} ]
% \end{equation}
% and can be estimated from samples, in particular by reusing the derivatives $\nabla_\bfpsi \log q_\bfpsi(\bftheta)$ already
% evaluated to form $\nabla_\bfpsi U(\bfpsi)$.

@@

    \label{alg:avo}
    \begin{algorithmic}[1]
        \State{$q(\bftheta|\bfpsi) \leftarrow \text{prior on $\bftheta$ (with differentiable and known density)}$}
        \While{$\bfpsi$ has not converged}
            \For{$i=1$ to $n_{\text{critic}}$} \Comment{Update $d$}
                \State{Sample a mini-batch $\{\mathbf{x}_m \sim p_r(\mathbf{x}), \bftheta_m \sim q(\bftheta|\bfpsi), \bfz_m \sim p(\bfz|\bftheta_m), \epsilon_m \sim U[0,1] \}_{m=1}^M$.}
                \For{$m=1$ to $M$}
                    \State{$\tilde{\mathbf{x}}_m \leftarrow g(\bfz_m; \bftheta_m)$}
                    \State{$\hat{\mathbf{x}}_m \leftarrow \epsilon_m \mathbf{x}_m + (1 - \epsilon_m) \tilde{\mathbf{x}}_m$}
                    \State{$U_d^{(m)} \leftarrow d(\tilde{\mathbf{x}}_m; \phi) - d(\mathbf{x}_m; \phi) + \lambda (|| \nabla_{\hat{\mathbf{x}}_m} d(\hat{\mathbf{x}}_m; \phi) ||_2 - 1)^2 $}
                \EndFor
                \State{$\phi \leftarrow \text{Adam}(\nabla_\phi \tfrac{1}{M} \sum_{m=1}^M U_d^{(m)})$}
            \EndFor
            \State{Sample a mini-batch $\{ \bftheta_m \sim q(\bftheta|\bfpsi),  \bfz_m \sim p(\bfz|\bftheta_m) \}_{m=1}^M$.}  \Comment{Update $q(\bftheta|\bfpsi)$}
            \State{$\nabla_\bfpsi U_g \leftarrow \tfrac{1}{M} \sum_{m=1}^M -d(g(\bfz_m; \bftheta_m)) \nabla_\bfpsi \log q_\bfpsi(\bftheta_m)$}
            % \State{$\mathbf{F}(\bfpsi) \leftarrow \tfrac{1}{M} \sum_{m=1}^M  \nabla_\bfpsi \log q_\bfpsi(\bftheta_m) \nabla_\bfpsi \log q_\bfpsi(\bftheta_m)^{\top}$}
            \State{$\nabla_\bfpsi H(q_\bfpsi) \leftarrow \tfrac{1}{M} \sum_{m=1}^M  \nabla_\bfpsi q_\bfpsi(\bftheta_m) \log q_\bfpsi(\bftheta_m)$}
            \State{$\bfpsi \leftarrow \text{Adam}(\nabla_\bfpsi U_g + \gamma \nabla_\bfpsi H(q_\bfpsi))$}
        \EndWhile
    \end{algorithmic}
    \end{algorithm}
    \end{minipage}
\end{figure*}

The alternating stochastic gradient descent on ${\cal L}_d$ and ${\cal L}_g$ in
GANs (Section~\ref{sec:gans}) inherently assumes that the generator $g$ is a differentiable function. In
the setting where we are interested in the inference of the parameters of a
fixed non-differentiable simulator (Section~\ref{sec:problem}),
rather than in learning the generative model itself,
gradients $\nabla_\bftheta g$ either do not exist or cannot be accessed. As a
result, gradients $\nabla_\bftheta {\cal L}_g$ cannot be constructed and the
optimization procedure cannot be carried out.

In this work, we propose to rely on variational optimization to minimize ${\cal
L}_d$ and ${\cal L}_g$, thereby bypassing the non-differentiability of $g$. More
specifically, we consider a proposal distribution $q(\bftheta|\bfpsi)$ over the
parameters of $g$ and $p(\mathbf{x}|\bftheta)$ and minimize in alternation the variational upper bounds
\begin{align}
    U_d &= \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [ {\cal L}_d ] \label{eqn:vo-ud} \\
    U_g &= \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [ {\cal L}_g ] \label{eqn:vo-ug}
\end{align} respectively over $\phi$ and $\bfpsi$.
When updating
$\phi$, unbiased estimates of $\nabla_\phi U_d$ can be obtained by
evaluating the exact and known gradient of $U_d$ over mini-batches of true and
generated data, as ordinarily done in stochastic gradient descent. When updating
$\bfpsi$, estimates of $\nabla_\bfpsi U_g$ can be derived with forward
simulations, as described in the previous section.
That is,
\begin{equation}\label{eqn:grad-ug-approx}
    \nabla_\bfpsi U_g = \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi), \bfz \sim p(\bfz|\bftheta)}  [-d(g(\bfz;\bftheta);\phi) \nabla_\bfpsi \log q(\bftheta|\bfpsi)],
\end{equation} which we can approximate with mini-batches of
generated data
\begin{equation}
    \nabla_\bfpsi U_g \approx \frac{1}{M} \sum_{m=1}^M -d(g(\bfz_m; \bftheta_m); \phi) \nabla_\bfpsi \log q(\bftheta_m|\bfpsi)
\end{equation}
for $\bftheta_m \sim q(\bftheta|\bfpsi)$ and $\bfz_m \sim p(\bfz|\bftheta_m)$.
For completeness, Algorithm~\ref{alg:avo} outlines the proposed adversarial variational
optimization (AVO) procedure, as built on top of WGAN-GP.
Obviously, the variational relaxation could similarly be coupled with
other variants of GANs and/or of evolution strategies.

Practically, the variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
a distribution parameterized in terms of $\bfpsi$:
\begin{equation}\label{eqn:p_psi}
    \mathbf{x} \sim \qxpsi \equiv \bftheta \sim q(\bftheta|\bfpsi), \bfz \sim p(\bfz|\bftheta), \mathbf{x} = g(\bfz; \bftheta).
\end{equation}
Intuitively, this corresponds to a family of simulators, each configured
with randomly sampled parameters $\bftheta \sim q(\bftheta|\bfpsi)$, whose joint collection
of generated samples is optimized with adversarial training to approach the real data distribution $p_r(\mathbf{x})$.
More formally, the learned model  therefore corresponds to the marginal distribution
of the generated data. That is,
\begin{equation}
    \qxpsi = \int  p(\mathbf{x}|\bftheta) q(\bftheta|\bfpsi) d\bftheta.
\end{equation}

In consequence, the proposed inference algorithm does not necessarily guarantee that the
proposal distribution $q(\bftheta|\bfpsi)$ will place its mass arbitrarily tight
around the parameters of interest, which might be an issue when one is rather interested in point estimates $\bftheta^*$.
For this purpose, we augment Eqn.~\ref{eqn:vo-ug}
with a regularization term corresponding to the differential entropy $H$ of
the proposal distribution. That is,
\begin{equation}
    U_g = \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [ {\cal L}_g ] + \gamma H(q(\bftheta|\bfpsi))
\end{equation}
where $\gamma \in \mathbb{R}^+$ is a hyper-parameter controlling the trade-off
between the generator objective and the tightness of the proposal distribution.
For small values of $\gamma$,
proposal distributions with large entropy are not penalized, which may result
in learning a smeared variation of the original simulator. On the other hand,
for large values of $\gamma$, the procedure is constrained to fit a proposal
distribution with low entropy, which has the effect of concentrating its density
tightly around one or a few $\bftheta$ values. A too large penalty may however
eventually make the optimization unstable, as the variance of $\nabla_\bfpsi \log q(\bftheta_m|\bfpsi)$
typically increases as the entropy of the proposal decreases.

\begin{eqnarray}
\lambda^* = \inf_\lambda \sup_\phi \mathbb{E}_{ z \sim q( z | \lambda)} [ (O^{p,q} f_\lambda) ] 
\end{eqnarray}
with
\begin{equation}
(O^{p,q} f) = \log q(z) - \log p(z|x) \, \forall f \in \mathcal{F}
\end{equation}

%This is natural here, but not quite OPVI formalism:
%\begin{equation}
%\bfpsi^* = \inf_\bfpsi \sup_\phi \mathbb{E}_{ \bftheta \sim q(\bftheta | \bfpsi)} [ (O^{p,q} f_\phi) ]
%\end{equation}
%
%\begin{equation}
%(O^{p,q} f) = W(p,q) = \sup_f \mathbb{E}_{x\sim p}[f(x)] - \mathbb{E}_{x \sim q}[f(x)]
%\end{equation}
%The expectation should be $x \sim p(x|\bfpsi)$ and here $(O^{p,q} f)$ is a scalar and not a function of $x$

>>>>

\newcommand{\qxpsi}{q(\mathbf{x}|\bfpsi)}
\newcommand{\bftheta}{{\bm \theta}}
\newcommand{\bfpsi}{{\bm \psi}}
\newcommand{\bfphi}{{\bm \phi}}
\newcommand{\bfx}{\mathbf{x}}
\newcommand{\bfz}{\mathbf{z}}

@@

Given some observed data $\{ \mathbf{x}_i | i=1, \dots, N \}$ drawn from the
(unknown) true distribution $p_r(\mathbf{x})$, our goal is to estimate the parameters 
$\bftheta^*$ that minimize the divergence between $p_r(\mathbf{x})$ and
the implicit model $p(\mathbf{x}|\bftheta)$. That is,
\begin{equation}
    \bftheta^* = \arg \min_{\bftheta \in \bftheta} \rho(p_r(\mathbf{x}), p(\mathbf{x}|\bftheta)),
\end{equation}
where $\rho$ is some distance or divergence. For many choices of divergence $\rho$, this is asymptotically equivalent to a maximum likelihood estimate.

@@

Generative adversarial networks (GANs) were first proposed by
\cite{goodfellow2014generative} as a way to build an implicit generative model
capable of producing samples from random noise $\bfz$. More specifically,
a generative model $g(\cdot; \bftheta)$ is pit against an adversarial
classifier $d(\cdot; \bfphi):\mathbb{R}^d \to [0,1]$ with parameters $\bfphi$ and whose antagonistic objective is to recognize real data $\mathbf{x}$
from generated data $\tilde{\mathbf{x}} = g(\bfz; \bftheta)$. Both models $g$ and $d$
are trained simultaneously, in such a way that $g$ learns to fool
its adversary $d$ (which happens when $g$ produces samples comparable to the
observed data), while $d$ continuously adapts to changes in $g$. When $d$ is
trained to optimality before each parameter update of the generator, it can
be shown that the original adversarial learning procedure~\cite{goodfellow2014generative} amounts to minimizing
the Jensen-Shannon divergence $\text{JSD}(p_r(\mathbf{x}) \parallel p(\mathbf{x}|\bftheta))$ between $p_r(\mathbf{x})$ and $p(\mathbf{x}|\bftheta)$.

As thoroughly explored in \citep{2017arXiv170104862A}, GANs remain remarkably
difficult to train because of vanishing gradients as $d$ saturates, or because of
unreliable updates when the training procedure is relaxed. As a remedy,
Wasserstein GANs~\citep{2017arXiv170107875A} reformulate the adversarial
setup in order to minimize the Wasserstein-1 distance $W(p_r(\mathbf{x}), p(\mathbf{x}|\bftheta))$ by
replacing the adversarial classifier with a 1-Lipschitz adversarial critic
$d(\cdot; \bfphi) : \mathbb{R}^d \to \mathbb{R}$. 
\begin{align}
W( p_r(\mathbf{x} ), p(\mathbf{x} | \bftheta)  ) = \sup_{d} \mathbb{E}_{p(\tilde{\mathbf{x}}|\bftheta)} [d(\tilde{\mathbf{x}})] - \mathbb{E}_{p_r(\mathbf{x})} [d(\mathbf{x})]
\end{align}
Under the WGAN-GP formulation of \cite{2017arXiv170400028G}
for stabilizing the optimization procedure,
training $d$ and $g$ results in alternating gradient updates on $\bfphi$ and $\bftheta$ in order to respectively minimize
\begin{align}
    {\cal L}_d =\,&  W(p_r, p_\theta) + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\bfphi) ||_2 - 1)^2] \\
    {\cal L}_g =\,& - W(p_r, p_\theta)
\end{align}
%
%\begin{align}
%    {\cal L}_d =\,&  {\cal L}_W + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\bfphi) ||_2 - 1)^2] \\
%    {\cal L}_g =\,& - {\cal L}_W
%\end{align}
%\begin{align}
%    {\cal L}_d =\,& \mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)} [d(\tilde{\mathbf{x}};\bfphi)] - \mathbb{E}_{\mathbf{x} \sim p_r(\mathbf{x})} [d(\mathbf{x};\bfphi)]  \nonumber \\
%                  & + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\bfphi) ||_2 - 1)^2] \\
%    {\cal L}_g =\,& -\mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)} [d(\tilde{\mathbf{x}};\bfphi)]
%\end{align}
where ${\hat{\mathbf{x}}} := \epsilon \mathbf{x} +
(1-\epsilon)\tilde{\mathbf{x}}$, for $\epsilon \sim U[0,1]$, $\mathbf{x} \sim
p_r(\mathbf{x})$ and $\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)$.

@@

Variational optimization~\cite{2012arXiv1212.4507S,staines2013optimization} and evolution strategies~\citep{2011arXiv1106.4487W} are general
optimization techniques that can be used to form a differentiable bound
on the optima of a non-differentiable function. Given a function $f$ to minimize,
these techniques are based on the simple fact that
\begin{equation}
    \min_{\bftheta \in {\Theta}} f(\bftheta) \leq \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta)] = U(\bfpsi),
\end{equation}
where $q(\bftheta|\bfpsi)$ is a proposal distribution with parameters $\bfpsi$ over input values $\bftheta$.
That is, the minimum of a set of function values is always less than or equal
to any of their average. Provided that the proposal is flexible enough, the parameters $\bfpsi$
can be updated to place its mass arbitrarily tight around the optimum $\bftheta^* = \min_{\bftheta \in \Theta} f(\bftheta)$.

Under mild restrictions outlined in  \citep{2012arXiv1212.4507S}, the bound
$U(\bfpsi)$ is differentiable with respect to $\bfpsi$, and using the log-likelihood
trick it can be rewritten as:
\begin{align}\label{eqn:approx-grad}
    \nabla_\bfpsi U(\bfpsi) &= \nabla_\bfpsi \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta)] \nonumber \\
    %&= \nabla_\bfpsi \int f(\bftheta)  q(\bftheta|\bfpsi)  d\bftheta \nonumber \\
 %   &= \int f(\bftheta) \nabla_\bfpsi q(\bftheta|\bfpsi)  d\bftheta \nonumber \\
   % &= \int \left[ f(\bftheta) \nabla_\bfpsi \log q(\bftheta|\bfpsi) \right]  q(\bftheta|\bfpsi) d\bftheta \nonumber \\
    &= \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta) \nabla_\bfpsi \log q(\bftheta|\bfpsi)]
\end{align}
Effectively, this means that provided that the score function $\nabla_\bfpsi \log
q(\bftheta|\bfpsi)$ of the proposal is known and that one can evaluate
$f(\mathbf{\bftheta})$ for any $\bftheta$, then one can construct empirical
estimates of Eqn.~\ref{eqn:approx-grad}, which can in turn be used to minimize
$U(\bfpsi)$ with stochastic gradient descent (or a variant thereof, like
Adam~\cite{2014arXiv1412.6980K} or the Natural Evolution Strategy
algorithm~\citep{2011arXiv1106.4487W}, for scaling invariance and
robustness to noisy gradients).

% In this simple form, stochastic gradient descent based on
% Eqn~\ref{eqn:approx-grad} is known to suffer from premature convergence and lack
% of scale invariance. To avoid the dependence on the parameterization of the
% proposal distribution, natural gradients~\citep{amari1998natural} can be used
% instead to point towards the steepest descent in the space of realizable
% proposal distributions, rather than in the space of their parameters. Effectively,
% the natural evolution strategy (NES) algorithm~\citep{2011arXiv1106.4487W}
% corrects the gradient of $U$ according to the local curvature of the KL-divergence
% surface of proposal distributions, which results in forming
% \begin{equation}\label{eqn:approx-grad-nat}
%     \tilde{\nabla}_\bfpsi U(\bfpsi) = \mathbf{F}(\bfpsi)^{-1} \nabla_\bfpsi U(\bfpsi),
% \end{equation}
% where $\mathbf{F}(\bfpsi)$ is the Fisher information matrix
% \begin{equation}
%     \mathbf{F}(\bfpsi) = \mathbb{E}_{\bftheta \sim q_\bfphi}[ \nabla_\bfpsi \log q_\bfpsi(\bftheta) \nabla_\bfpsi \log q_\bfpsi(\bftheta)^{\top} ]
% \end{equation}
% and can be estimated from samples, in particular by reusing the derivatives $\nabla_\bfpsi \log q_\bfpsi(\bftheta)$ already
% evaluated to form $\nabla_\bfpsi U(\bfpsi)$.

@@

    \label{alg:avo}
    \begin{algorithmic}[1]
        \State{$q(\bftheta|\bfpsi) \leftarrow \text{prior on $\bftheta$ (with differentiable and known density)}$}
        \While{$\bfpsi$ has not converged}
            \For{$i=1$ to $n_{\text{critic}}$} \Comment{Update $d$}
                \State{Sample a mini-batch $\{\mathbf{x}_m \sim p_r(\mathbf{x}), \bftheta_m \sim q(\bftheta|\bfpsi), \bfz_m \sim p(\bfz|\bftheta_m), \epsilon_m \sim U[0,1] \}_{m=1}^M$.}
                \For{$m=1$ to $M$}
                    \State{$\tilde{\mathbf{x}}_m \leftarrow g(\bfz_m; \bftheta_m)$}
                    \State{$\hat{\mathbf{x}}_m \leftarrow \epsilon_m \mathbf{x}_m + (1 - \epsilon_m) \tilde{\mathbf{x}}_m$}
                    \State{$U_d^{(m)} \leftarrow d(\tilde{\mathbf{x}}_m; \bfphi) - d(\mathbf{x}_m; \bfphi) + \lambda (|| \nabla_{\hat{\mathbf{x}}_m} d(\hat{\mathbf{x}}_m; \bfphi) ||_2 - 1)^2 $}
                \EndFor
                \State{$\bfphi \leftarrow \text{Adam}(\nabla_\bfphi \tfrac{1}{M} \sum_{m=1}^M U_d^{(m)})$}
            \EndFor
            \State{Sample a mini-batch $\{ \bftheta_m \sim q(\bftheta|\bfpsi),  \bfz_m \sim p(\bfz|\bftheta_m) \}_{m=1}^M$.}  \Comment{Update $q(\bftheta|\bfpsi)$}
            \State{$\nabla_\bfpsi U_g \leftarrow \tfrac{1}{M} \sum_{m=1}^M -d(g(\bfz_m; \bftheta_m)) \nabla_\bfpsi \log q_\bfpsi(\bftheta_m)$}
            % \State{$\mathbf{F}(\bfpsi) \leftarrow \tfrac{1}{M} \sum_{m=1}^M  \nabla_\bfpsi \log q_\bfpsi(\bftheta_m) \nabla_\bfpsi \log q_\bfpsi(\bftheta_m)^{\top}$}
            \State{$\nabla_\bfpsi H(q_\bfpsi) \leftarrow \tfrac{1}{M} \sum_{m=1}^M  \nabla_\bfpsi q_\bfpsi(\bftheta_m) \log q_\bfpsi(\bftheta_m)$}
            \State{$\bfpsi \leftarrow \text{Adam}(\nabla_\bfpsi U_g + \gamma \nabla_\bfpsi H(q_\bfpsi))$}
        \EndWhile
    \end{algorithmic}
    \end{algorithm}
    \end{minipage}
\end{figure*}

The alternating stochastic gradient descent on ${\cal L}_d$ and ${\cal L}_g$ in
GANs (Section~\ref{sec:gans}) inherently assumes that the generator $g$ is a differentiable function. In
the setting where we are interested in estimating the parameters of a
fixed non-differentiable simulator (Section~\ref{sec:problem}),
rather than in learning the generative model itself,
gradients $\nabla_\bftheta g$ either do not exist or are inaccessible. As a
result, gradients $\nabla_\bftheta {\cal L}_g$ cannot be constructed and the
optimization procedure cannot be carried out.

In this work, we propose to rely on variational optimization to minimize ${\cal
L}_d$ and ${\cal L}_g$, thereby bypassing the non-differentiability of $g$. More
specifically, we consider a proposal distribution $q(\bftheta|\bfpsi)$ over the
parameters of $g$ and $p(\mathbf{x}|\bftheta)$ and alternately minimize the variational upper bounds
\begin{align}
    U_d &= \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [ {\cal L}_d ] \label{eqn:vo-ud} \\
    U_g &= \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [ {\cal L}_g ] \label{eqn:vo-ug}
\end{align} respectively over $\bfphi$ and $\bfpsi$.
When updating
$\bfphi$, unbiased estimates of $\nabla_\bfphi U_d$ can be obtained by
directly evaluating the gradient of $U_d$ over mini-batches of real and
generated data, as is ordinarily done in stochastic gradient descent. When updating
$\bfpsi$, $\nabla_\bfpsi U_g$ can be estimated as described in the previous section.
That is,
%\begin{equation}\label{eqn:grad-ug-approx}
%   \nabla_\bfpsi U_g = \mathbb{E}_{\subarray{l}{\bftheta \sim q(\bftheta|\bfpsi), \\ \hat{\bfx} \sim p(\bfx | \bftheta) }}  [-d( \hat{\bfx} ;\bfphi) \nabla_\bfpsi \log q(\bftheta|\bfpsi)],
%\end{equation} 
\begin{equation}\label{eqn:grad-ug-approx}
   \nabla_\bfpsi U_g = \mathbb{E}_{\substack{\bftheta \sim q(\bftheta|\bfpsi), \\ \hat{\bfx} \sim p(\bfx | \bftheta)}}  [-d( \hat{\bfx} ;\bfphi) \nabla_\bfpsi \log q(\bftheta|\bfpsi)],
\end{equation} 
%\begin{equation}\label{eqn:grad-ug-approx}
%    \nabla_\bfpsi U_g = \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi), \bfz \sim p(\bfz|\bftheta)}  [-d(g(\bfz;\bftheta);\bfphi) \nabla_\bfpsi \log q(\bftheta|\bfpsi)],
%\end{equation} 
which we can approximate with mini-batches of
generated data
\begin{equation}
    \nabla_\bfpsi U_g \approx \frac{1}{M} \sum_{m=1}^M -d(g(\bfz_m; \bftheta_m); \bfphi) \nabla_\bfpsi \log q(\bftheta_m|\bfpsi)
\end{equation}
for $\bftheta_m \sim q(\bftheta|\bfpsi)$ and $\bfz_m \sim p(\bfz|\bftheta_m)$.
For completeness, Algorithm~\ref{alg:avo} outlines the proposed adversarial variational
optimization (AVO) procedure, as built on top of WGAN-GP.
Obviously, the variational relaxation could similarly be coupled with
other variants of GANs and/or of evolution strategies.

%Practically, the variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
%have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
%a distribution parameterized in terms of $\bfpsi$:
%\begin{equation}\label{eqn:p_psi}
%    \mathbf{x} \sim \qxpsi \equiv \bftheta \sim q(\bftheta|\bfpsi), \bfz \sim p(\bfz|\bftheta), \mathbf{x} = g(\bfz; \bftheta).
%\end{equation}
%Intuitively, this corresponds to a family of simulators, each configured
%with randomly sampled parameters $\bftheta \sim q(\bftheta|\bfpsi)$, whose joint collection
%of generated samples is optimized with adversarial training to approach the real data distribution $p_r(\mathbf{x})$.
%More formally, the learned model  therefore corresponds to the marginal distribution
%of the generated data. That is,
%\begin{equation}
%    \qxpsi = \int  p(\mathbf{x}|\bftheta) q(\bftheta|\bfpsi) d\bftheta.
%\end{equation}
%Intuitively, this corresponds to a family of simulators, each configured
%with randomly sampled parameters $\bftheta \sim q(\bftheta|\bfpsi)$, whose joint collection
%of generated samples is optimized with adversarial training to approach the real data distribution $p_r(\mathbf{x})$.

The variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
the marginal distribution parameterized by $\bfpsi$
\begin{equation}
    \qxpsi = \int  p(\mathbf{x}|\bftheta) q(\bftheta|\bfpsi) d\bftheta.
\end{equation}
Operationally, we sample from this marginal model via
\begin{equation}\label{eqn:p_psi}
    \mathbf{x} \sim \qxpsi \equiv \bftheta \sim q(\bftheta|\bfpsi), \bfz \sim p(\bfz|\bftheta), \mathbf{x} = g(\bfz; \bftheta).
\end{equation}
Effectively, we are optimizing a variational posterior $q(\bftheta | \psi)$ by minimizing the 
divergence of the corresponding posterior predictive distribution $\qxpsi$ and the data distribution $p_r(\bfx)$. 
%Effectively, we are performing variational inference through model criticism~\citep{box1980sampling}.
Consequently, the proposed algorithm is a form of variational Bayesian inference through model criticism~\citep{box1980sampling}.

%Consequently, the proposed inference algorithm does not necessarily guarantee that the
%proposal distribution $q(\bftheta|\bfpsi)$ will place its mass arbitrarily tight
%around the parameters of interest, which might be an issue when one is rather interested in point estimates $\bftheta^*$.
%For this purpose, we augment Eqn.~\ref{eqn:vo-ug}
